---
title: "Environmental Analysis"
author: "Rhiannon Nolan"
date: "April 02, 2023"
output:
  html_document:
  toc: true
editor_options: 
  chunk_output_type: console
---

```{R, echo=FALSE}
#GLOBAL R chunk options.
#   (to hide this message add "echo=FALSE" to the code chunk options)

knitr::opts_chunk$set(comment = NA, message = FALSE, warning = FALSE, width = 100)
knitr::opts_chunk$set(fig.align = "center", fig.height = 4, fig.width = 6)

#knitr::opts_chunk$set(cache = TRUE, autodep=TRUE)

```

This script uses csv files created after maxent models have been run for all species across all time bins. 
First the tables of results are filtered to include only environmental pixels where threshold = 1 (inside the realized niche).

## Set Up Workspace

```{r}
#install necessary packages
if (!require(rstudioapi)){install.packages("rstudioapi")}
me <- (rstudioapi::getActiveDocumentContext()$path) #get current file name

if (!require(renv)){ install.packages("renv")}
packs <- unique(renv::dependencies(me)$Package) #list of packages needed 

for (i in seq_along(packs)){
  if (!require(packs[i], character.only=T)){ #install any missing pacls
    install.packages(packs[i])
  }
}

lapply(packs, library, character.only=T) #load packages
rm(me, packs, i) #clean env
```


## Load in all csv files
```{R}
setwd(dirname(rstudioapi::getSourceEditorContext()$path)) #set file location as wd


file <- list.files(paste0(getwd()), pattern = "csv", full.names = TRUE) #find csv file with data

#choose which one is actually right
file <- file[4]

data <- read.csv(file) #make data frame of imported data
data <- tibble::rowid_to_column(data, var = "ID") #add row ID column
data <- data %>% subset(
  select = c(ID, Name, long, lat, time.bin, suitability, mahal)
)
head(data) #does it look right?


data$times <- data$time.bin #make a new variable that stays a character for simplicity later on
unique(data$times) #list all time bins present

#change time bin order for plotting
data$time.bin <- as.factor(data$time.bin)
levels(data$time.bin) #what is the current order?

#change order or components if necessary
data$time.bin <- factor(data$time.bin, levels=c('EPleist', 'MPleist', 'LPleist', 'Holo', 'Modern', 'Cumul'))


#filter to include the expanded training region only for a "realized niche" model
achenm.occ <- read.csv("D:/202508_MaxentModels/3_MaxentModels_OldData/master.occs.csv", header=T)
min.lng <- min(achenm.occ$long) - 2
max.lng <- max(achenm.occ$long) + 2
min.lat <- min(achenm.occ$lat) - 2
max.lat <- max(achenm.occ$lat) + 2

data_RN <- data %>% filter(
    long >= min.lng & long <= max.lng & 
    lat >= min.lat & lat <= max.lat)

#optional
data <- data_RN

try(rm(file, min.lng, max.lng, min.lat, max.lat, achenm.occ, data_RN))
```

## Set up Datasets

```{r}
#set up data groupings by species and time bin
list_of_data <- list()
times <- unique(data$time.bin)

for (time in times) { #divide data into time bins
  dat <- filter(data, time.bin == time)
  spp <- as.list(unique(dat$Name))
  
  for (i in seq_along(spp)) { #divide data into species
    data_subset <- as.data.frame(filter(dat, Name == spp[[i]]))
    name <- stringr::str_split_1(spp[[i]], "_") #set up sample name
    sample <- paste(substr(name[1], 1, 1), substr(name[-1], 1, 3), sep = ".")
    sample <- paste(sample, time, sep = ".") #set up name
    list_of_data[[sample]] <- data_subset
  }
  }

rm(dat, spp, data_subset, name, sample, time, i, times) #clean env

#verify the list of data looks right
names(list_of_data) #datasets in list
head(list_of_data[[2]]) #see which species is which, change number to change species
```

```{r}
#set up bins of only suitable data - where habitability score > threshold
bins_hist <- list() #make empty list of histograms

for(i in seq_along(list_of_data)){
  dat <- as.data.frame(list_of_data[[i]])
  dat <- filter(dat, suitability > 0) #filter to suitable data
  if(nrow(dat)>0){ #perform only if there is data at that time
    name <- names(list_of_data)[i] #keep names consistent
    bins_hist[[name]] <- dat
    rm(dat, name)
  } else {
  print(paste0("No data for ", dat$Name[1], " at time ", i)) #lets us know there were empty times
  }
}
rm(i) #clean env
```
###Hereafter, list_of_data includes all data in each bin, bins_hist includes only suitable data.

## Linear Models

```{r}
#make linear models

histograms <- list()
sequence_data <- list()

#set up histograms
for (i in seq_along(bins_hist)) {
  #make histogram
  summ <- summary(bins_hist[[i]]$mahal) #set up summary object (to get min mahal dist max&min)
  species <- bins_hist[[i]]$Name[1] #name the histogram
  time <- bins_hist[[i]]$time.bin[1] #time period
  run <- names(bins_hist)[i]
  hist_mahal <- hist(bins_hist[[i]]$mahal, 
                     breaks = nrow(subset(bins_hist[[i]], time.bin == time)), 
                     main = paste0("Mahalanobis frequencies for ", species, " during the ", time))
  histograms[[run]] <- hist_mahal #save histogram
  
  #get mahalanobis distances and frequencies for lms
  sequence <- data.frame(
    species = species,
    time = time,
    id = seq_along(hist_mahal$counts), # Create an ID for bins
    mahal = seq(from = summ[[1]], #for more accurate information, use mahal dist values
              to = summ[[6]],
              by = (summ[[6]]-summ[[1]])/(length(hist_mahal$counts)-1)),
    freq = hist_mahal$counts          # Use the counts from the histogram
  )
  rows <- round((nrow(sequence)*0.97), digits = 0) 
  sequence <- sequence %>% filter(row_number() <= n()-rows) #filter out the last 3% (use only 97% of suit hab)
  
  sequence_data[[run]] <- sequence
}

try(rm(species, time, i, rows, run, summ, hist_mahal, sequence))
```

```{r}
#set up to get p-value from lms
p_val <- function(model) {
    f <- summary(model)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}


#make linear models using the histogram data
list_of_lms <- list()
lm_coefs<- data.frame()
columns= c("species","time","R2","Adj_R2","p-value") 
lm_R2p <- data.frame()
plots <- list()
seg_fits <- list()
segfits_data <- list()
seg_plots <- list()


for (i in seq_along(sequence_data)){
  run <- names(sequence_data)[i]
  species <- sequence_data[[i]]$species[1] #name the histogram
  time <- as.character(sequence_data[[i]]$time[[1]]) #time period
  
  #fit linear models
  freqmahal <- lm(freq ~ mahal, data = sequence_data[[i]])
  list_of_lms[[run]] <- freqmahal
  
  #segmented fit
  seg <- segmented::segmented(freqmahal, seg.Z = ~ mahal)
  seg_fits[[run]] <- seg
  fit <- numeric(length(sequence_data[[i]]$mahal))
  fit[complete.cases(rowSums(cbind(sequence_data[[i]]$freq, sequence_data[[i]]$mahal)))] <- broken.line(seg)$fit
  dat_segfit <- data.frame(mahal = sequence_data[[i]]$mahal, freq = sequence_data[[i]]$freq, fit = fit)
  segfits_data[[run]] <- dat_segfit
  
  
  #get coefficients, R2 and p-value
  R2p <- c(species, time, summary(freqmahal)[[8]], summary(freqmahal)[[9]], p_val(freqmahal))
  lm_R2p <- rbind(lm_R2p, R2p)
  
  coeffs <- as.data.frame(summary(freqmahal)$coefficients, col.names = T, row.names = F)
  coeffs$Value <- c("Intercept", "Mahal")
  coeffs$time <- time
  coeffs$species <- species
  lm_coefs <- rbind(lm_coefs, coeffs)
  
  #title <- paste0(species, " at time ", time) 
  #make plots
  plot <- ggplot2::ggplot(sequence_data[[i]], aes(x = mahal, y = freq)) +
    geom_point(color = "#74A089") +
    geom_smooth(method = "lm", color = "#CB7A5C") +
    
    labs(x = "Mahal dist", y = "Freq (~Abun)") +
    annotate("text", x = min(sequence_data[[i]]$mahal), y = max(sequence_data[[i]]$freq), 
             label = paste("R² =", round(as.numeric(R2p[4]), 3)), hjust = 0, vjust = 1, size = 3, color = "darkgray") +
    theme_minimal() +
    ylim(0, NA)  # Set lower limit to 0 and keep upper limit as it is
  
  # Save the plot to a variable with a dynamic name
  plots[[run]] <- plot
  rm(plot)
  
  #make segmented fit plots
  plot <- ggplot() +
        geom_point(data = dat_segfit, 
                   aes(x = mahal, y = freq), color = "#525174") +
        geom_line(data = dat_segfit, aes(x = mahal, y = fit), color = "#68B0AB", linewidth = 1) +
        labs(x = "Mahal dist", y = "Freq (~Abun)") +
        annotate("text", x = min(sequence_data[[i]]$mahal), y = max(sequence_data[[i]]$freq), 
             label = paste("R² =", round(as.numeric(summary(seg)[[9]]), digits = 3)), hjust = 0, vjust = 1, size = 3, color = "darkgray") +
        theme_minimal() +
        ylim(0, NA)  # Set lower limit to 0 and keep upper limit as it is
  
  seg_plots[[run]] <- plot
}

#fix names in the R2 and p-value df
colnames(lm_R2p) = columns

options(suppressWarnings(try(rm(i, summ, species, time, run, hist_mahal, 
                                sequence, rows, columns, R2p, p_val, coeffs, plot,
                                dat_segfit, freqmahal, lm_coefs, lm_R2p, seg, fit)))) #clean env
```

```{r}
#compare habitat suitability and mahalanobis distance to verify base assumption of ACH
suitfits <- list()
suitplots <- list()


for (i in seq_along(list_of_data)){
  run <- names(list_of_data)[i]
  
    #fit linear models
  lm <- lm(suitability ~ mahal, data = list_of_data[[i]])
  suitfits[[run]] <- lm
  
  #make plots
    #make plots
  plot <- ggplot(list_of_data[[i]], aes(x = mahal, y = suitability)) +
    geom_point(color = "#5E548E") +
    geom_smooth(method = "lm", color = "#3A506B") +
    
    labs(x = "Mahal dist", y = "Suitability") +
    annotate("text", x = min(list_of_data[[i]]$mahal), y = max(list_of_data[[i]]$suitability), 
             label = paste("R² =", round(as.numeric(summary(lm)[[9]]), 3)), hjust = 0, vjust = 1, size = 3, color = "black") +
    theme_minimal() +
    ylim(0, NA)  # Set lower limit to 0 and keep upper limit as it is
  
  suitplots[[run]] <- plot
  
}

try(rm(run, lm, plot, i,))
```


## Make Figures

```{r}
# Define the layout matrix for plots
layout_matrix <- rbind(
  # EP  MP  LP  H  M  C
  c(1,  21, 9   ,3  ,14, 24 ), #Aaeq
  c(2,  22, 10  ,4  ,15, 25 ), #Acor
  c(NA, 23, 11  ,5  ,16, 26 ), #Cele 
  c(NA, NA, 12  ,6  ,17, 27 ), #Lova
  c(NA, NA, 13  ,7  ,18, 28 ), #Mcam
  c(NA, NA, NA  ,8  ,19, 29 ), #Ptri
  c(NA, NA, NA, NA  ,20, 30 )  #Sflo
)

```

```{r}
#linear models mahal vs frequency
gridExtra::grid.arrange(grobs = plots,
             layout_matrix = layout_matrix)

#segmented models mahal vs frequency
gridExtra::grid.arrange(grobs = seg_plots,
             layout_matrix = layout_matrix)

#suitability vs mahalanobis dist
gridExtra::grid.arrange(grobs = suitplots,
             layout_matrix = layout_matrix)
```

## Spearman Correlations

```{r}
#mahalanobis distance vs habitat suitability
corrs_mahalsuit <- data.frame()

for (i in seq_along(list_of_data)){
  species <- list_of_data[[i]]$Name[1] #name the histogram
  time <- as.character(list_of_data[[i]]$time.bin[1]) #time period
  corr <- cor.test(as.numeric(list_of_data[[i]]$suitability), #set up spearman test
                   as.numeric(list_of_data[[i]]$mahal),
                   data = list_of_data[[i]],
                   method = 'spearman')
  if ( corr$p.value < 0.05) {
    results <- c(species, time, corr$estimate, corr$p.value)
  } else { 
      results <- c(species, time, NA, NA)}
  corrs_mahalsuit <- rbind(corrs_mahalsuit, results)
}

colnames(corrs_mahalsuit) = c("Species", "Time", "Rho", "p-value")

path <- "D:/202508_MaxentModels/6_RPlots/Spearman_MahalSuit.csv"
write.csv(corrs_mahalsuit, file = path)

try(rm(species, time, corr, results, path, i))
```

```{r}
#mahalanobis distance vs frequency
corrs_mahalfreq <- data.frame()

for (i in seq_along(sequence_data)){
  species <- sequence_data[[i]]$species[1] #name the histogram
  time <- as.character(sequence_data[[i]]$time[1]) #time period
  corr <- cor.test(as.numeric(sequence_data[[i]]$freq), #set up spearman test
                   as.numeric(sequence_data[[i]]$mahal),
                   data = sequence_data[[i]],
                   method = 'spearman')
  if ( corr$p.value < 0.05) {
    results <- c(species, time, corr$estimate, corr$p.value)
  } else { 
      results <- c(species, time, NA, NA)}
  corrs_mahalfreq <- rbind(corrs_mahalfreq, results)
}

colnames(corrs_mahalfreq) = c("Species", "Time", "Rho", "p-value")

path <- "D:/202508_MaxentModels/6_RPlots/Spearman_MahalFreq.csv"
write.csv(corrs_mahalfreq, file = path)

try(rm(species, time, corr, results, path, i))
```

